\chapter*{Acknowledgements}
Thanks go to my first supervisor Peer-Olaf Siebers, who always very patiently reminded me that a Ph.D. is not about to change the world but learning how to do research on my own. He was a strong guidance throughout my 3 years in Nottingham and I could not have hoped for a better and more dedicated first supervisor.

I am also thankful for my second supervisor Thorsten Altenkirch, who gave strong and sometimes brutal feedback about the technical details of my approaches. Due to the fact that his main interest is a rather theoretical spin on functional programming and computing, I am deeply grateful for his strong support of my rather practical approach to functional programming.

I am in depth to the whole Functional Programming Lab at UoN, for welcoming me in their midst despite my lack of specific theoretical background. I owe them many open and deep discussions which resulted in new insights. Further, presenting at their \textit{FP Lunch} was always a challenging but highly rewarding activity, always resulting in valuable feedback.

I am especially in depth to Ivan Perez for always having an open ear for questions and valuable discussions about his research, without this Ph.D. would have probably developed a very different spin.

Many thanks go to Martin Handley and James Hey for many discussions, feedback and proof reading of my papers and my thesis.

Thanks go also to Julie Greensmith for valuable discussions and pointing me into right directions at important stages of the Ph.D.