\chapter{Discussion}
\label{ch:discussion}
% starting point
This thesis started out by challenging the established views that \textit{"[..] object-oriented programming to be a particularly natural development environment for Sugarscape specifically and artificial societies generally [..]"} \cite{epstein_growing_1996} (p. 179) and that \textit{agents map naturally to objects} \cite{north_managing_2007}. As an alternative, a radical different approach to implementing ABS was proposed, using the pure functional programming paradigm. As language of choice, Haskell was motivated due to its  pure functional features, matureness and increasing relevance to real-world applications. 

% motivation
The conjecture was that by using Haskell, one can directly transfer the promises made by pure functional programming to ABS as well, directly gaining a few highly important benefits.  The relevance of each of these promises to ABS was already pointed out in the respective chapters and it is quite obvious that these benefits would clearly be of substantial value in ABS. The common baseline is that all those benefits support implementing ABS which are more likely to be correct, something of fundamental value in simulation.

\begin{enumerate}
	\item The static strong type system allows to remove a substantial number and class of bugs at run time and if one programs careful one can even guarantee that no bugs as in crashes or exceptions will occur at run time. This is particularly the case for purely computational problems, without \textit{IO} \footnote{Obviously IO is involved in all ABS, otherwise the results are not observable. By purely computational, we mean the lack of \textit{IO} in the agents, as fundamental part of the model specification, other than visualisation and exporting to an output file.}, as ABS almost always are. 
	
	\item Explicit handling and control of side effects delivers even more static guarantees at compile time and allows to deal with deterministic side effects (random-number streams, read-/write only contexts, state) in a referential transparent way. In combination with strong static typing this allows to reduce logical bugs, subject to the domain of the problem, by dramatically reducing available, valid operations on data - after all stateful applications are a fact, the challenge is how to deal with state. As ABS is an inherently stateful problem due to agents and the environment, this should increase the correctness of an ABS implementation as well. Further, this should allow to produce an implementation which is guaranteed to reproducible at compile time, where runs with same initial conditions \textit{will} result in same dynamics.
	
	\item Parallel and concurrent programming is claimed to be a lot easier, less painful and less error prone in functional programming in general and in Haskell in particular due to immutable data and the explicit handling of side effects. The concept of Software Transactional Memory, which allows to express a problem as a data-flow problem is highly promising. Besides, data-parallel programming promises to speedup code without the need for changing any of the logic or types. This seemed to offer a straightforward way of speeding up ABS implementations either through data parallelism or concurrency. This has always been quite difficult to achieve in traditional object-oriented ABS and pure functional programming seems to offer a solution.
	
	\item The data-centric declarative style, referential transparency and immutability of data makes testing substantially easier due to composability: functions can be easily tested in isolation from each other even if they involve side effects. This opens the door for randomised property-based testing which intuitively seemed to be a perfect match to test ABS implementations which are almost always stochastic in nature.
\end{enumerate}

The central question which needed to be answered first was \textit{how} ABS could be done pure functionally, as there didn't exist any research which offered a systematic solution to that problem. More specifically, it was unclear how to represent agents, how to express agent identity, local agents state, changing behaviour and interactions amongst agents and the environment. After all, this is straightforward in object-oriented programming due to method calls and mutable shared state encapsulated in objects. This thesis solution was to use arrowized FRP, both in the pure implementation of Yampa and the monadic version as in the library Dunai. Building on top of them allowed to implement pro-activity of agents, encapsulation of local agent state, an environment as shared mutable state and synchronous agent-interactions based on an event-driven approach. The central concept behind these approaches are Signal Functions (SF), generalised in Dunai to Monadic Stream Functions (MSF), which are implemented using closures and continuations, fundamental building blocks and concepts of pure functional programming. With theses techniques it became possible to implement time-driven models like the agent-based SIR as introduced in Chapter \ref{sec:sir_model} and the highly complex event-driven Sugarscape model as introduced in Chapter \ref{sec:sugarscape}. The fact that the developed concepts can manage the complexity of a full Sugarscape implementation is proof that they are suitable for most agent-based models, running on single machines.

The thesis then turned to the additional opportunities a pure functional approach would allow, with focus on parallelism and concurrency and property-based testing. By applying compositional parallelism it was possible to speed up the non-monadic time-driven ABS by a substantial factor, without losing static guarantees about reproducibility. The same was not possible with a monadic approach, which motivated the transition to concurrency using Software Transactional Memory, where a substantial speedup was achieved in a monadic, event-driven implementation sacrificing purity. Still, by limiting the side-effects to Software Transactional Memory only guarantees that the differences of the dynamics of repeated runs, come only from the side-effects of concurrency and nothing else.

The last step was to apply property-based testing to pure functional ABS, to test the hypothesis whether stochastic ABS and random property-based testing are a good match. Indeed, it has turned out to be a perfect match: it was possible to fully specify the agents behaviour of the time- and event-driven SIR model directly in code and verify them with QuickCheck. Further, other important techniques relevant for ABS like validation against specifications and comparisons of different implementations of the same model are easily done with property-based testing. Also, by showing how to perform hypothesis testing with property-based testing in the case of the Sugarscape model, we were able to show how to apply it to exploratory models which have no formal ground truth as well.

We argue that the thesis has successfully demonstrated in the respective chapters that all promises can successfully be transferred to pure functional ABS implementations in a reasonable, robust and maintainable way. Below we discuss respective points more in depth and give a holistic reflection and critique of our pure functional approach in general.

\input{./tex/discussion/benefits.tex}

\input{./tex/discussion/drawbacks.tex}

\input{./tex/discussion/gintisCase.tex}

\input{./tex/discussion/generalising.tex}

\input{./tex/discussion/whenNot.tex}

\input{./tex/discussion/objects.tex}