\chapter{Pure functional event-driven ABS}
\label{ch:eventdriven}
In this chapter we build on the previous discussion of update strategies in Chapter \ref{ch:impl_abs} and the implementation techniques presented in the time-driven approach of Chapter \ref{ch:timedriven} to develop concepts for event-driven ABS in a pure functional way. 

\medskip

In event-driven ABS \cite{meyer_event-driven_2014}, the simulation is advanced through events. Agents and the environment schedule events into the future and react to incoming events scheduled by themselves, other agents, the environment or the simulation kernel. Time is discrete in this approach and it advances step-wise from event to event, where each event has an associated receiver and $\Delta t$, indicating the delay to the current virtual simulation time when the event should be scheduled. This implies that time could stay constant, for example when an event is scheduled with $\Delta t = 0$ the virtual simulation time does not advance. Further, agents can schedule events to themselves, emulating a recurring behaviour, which in turn emulates pro-active behaviour. Because agents can adopt and change their state and behaviour when processing an event, this means that even if time does not advance, agents can change. This non-signal behaviour is the fundamental difference to the time-driven approach in Chapter \ref{ch:timedriven}. Further, this mechanism is used to implement synchronous agent interactions pure functionally as discussed in the respective sections below.

The event-driven approach makes the simulation kernel technically closely related to a Discrete Event Simulation \cite{zeigler_theory_2000}. Due to the necessity of imposing a correct ordering of events in this type of ABS, it needs to be stepped event by event, with the \textit{sequential} update strategy, as introduced in Chapter \ref{sec:seq_strategy}. It is important to emphasise that only the semantics of the sequential update strategy allow the kind of features  presented in the following sections, as the agents act one after the other, seeing the effects of previous agents in the same time step. This would not make sense in the parallel update strategy as used in time-driven ABS, where agents act conceptually at the same time. This means that event-driven ABS is inherently sequential due to its fundamental reliance on effects as will become clearer in the sections below. There exists also Parallel Discrete Event Simulation \cite{fujimoto_parallel_1990}, which processes events in parallel and deals with inconsistencies by reverting to consistent states. We hypothesise that a pure functional approach could be beneficial in such an approach due to persistent data structures and explicit handling of side effects but we leave this for further research.

\medskip

We start the chapter by introducing the concepts of agent identity and event scheduling using an event-driven agent-based SIR model, inspired by \cite{macal_agent-based_2010}. We then use the highly complex Sugarscape model as introduced in Chapter \ref{sec:sugarscape}, to develop more advanced features of ABS in a pure functional context: dynamic creation and removal of agents during simulation, adding a shared mutable environment, local mutable agent state and synchronous agent interactions. 
%Note that the Sugarscape model is not a real event-driven model like the event-driven SIR one is as in it the agents do schedule events but they don't do this into the future - events in Sugarscape don't have associated time-stamps.

\input{./tex/eventdriven/basics.tex}

\input{./tex/eventdriven/advanced.tex}

%\input{./tex/eventdriven/implementation.tex}

%\input{./tex/eventdriven/eventdrivenSIR.tex}

\input{./tex/eventdriven/discussion.tex}