\documentclass[a4paper, 10pt, conference]{../../templates/IEEEconf/IEEEconf}

\usepackage{graphicx}
\usepackage{caption} 
\usepackage{subcaption}
\usepackage{hyperref}
\usepackage{listings}
\usepackage{hhline}
\usepackage[utf8]{inputenc}
\usepackage{float}
\usepackage{amssymb}
\usepackage[autostyle=true]{csquotes}
\usepackage{amsmath}

\title{The Future of Agent-Based Simulation \\ Metaphysics and Genesis according to Computer Science}

\author{
	Jonathan Thaler \\
	\email{jonathan.thaler@nottingham.ac.uk} \\
	\begin{affiliation}
		School of Computer Science, University of Nottingham
	\end{affiliation} \\
}

\begin{document}
\maketitle

\begin{abstract}
As Agent-Based Simulation (ABS) is still a young discipline we argue that it has still a bright future ahead. We claim that its logical future is to simulate our own world, thus creating a point where simulation and reality converge, an event we postulate and term \textit{Simulation-Convergence}. Just by being able to talk about this convergence hints that there may be strong connections between the real world and ABS. As we will show the only really interesting and worthy reason for pursuing the \textit{Simulation-Convergence} is to simulate free will. We argue that the emergent properties of such a simulation of the free will are ideologies - mechanisms for restricting free will. We hypothesize that the \textit{Simulation-Convergence} may already have happened and that reality as we experience it is indeed an ABS created by our very self. This will give us a framework which allows us to talk about metaphysics from a computer-scientific view-point the first time. The remainder of the paper then interprets various metaphysical concepts like birth, death, karma, god, near-death experiences, afterlife, ethics, meditation, spiritual enlightenment, spiritual calling,... in this computer-scientific simulation-of-a-free-will context.


Main message: we could live in a simulation created by our selves.
	Hypothesis: 
		- It is a simulation of free will 
		- granting free will outside of a simulation would be too dangerous
		- emergent properties of free will are: ideologies which create inequality, suffering, death. 
		- the reason why these properties emerge is because the agents don't perceive the creation as one but as divided, thus leading to ideologies to explain the divisions thus leading to further division.
		- how could free will, conciousness and mind work in computation? hypothesis: all build on each other
		- hypothesis: the simulation was created by entities which hadnt free will. they split their conciousness into pieces (souls) and planted them into the simulation. this enabled them to learn.

[ ] we take free will for granted - it is not
[ ] in spirit there is not the kind of free will we have here. this cannot mean that all in spirit are robots, they just follow the rules because they agreed to: the give up free will willingly: they used free will to give it up (but its still there)

All sciences have their genesis-model which are basically explanations of how the world did come into existence, what the reason for existence is and who or what God is. Unfortunately computer science has none so far, so the aim of this paper is to set out to develop such genesis-model from a computer-science perspective. The model is motivated from the perspective that the world and humankind is a simulation to see free will in action in a sand-boxed environment as opposed to the afterlife / after-world or just the Beyond, which is an outer level of simulation and itself again a simulation as will be shown in subsequent sections. \\
This paper addresses important fundamental questions of belief and religion and tries to explain them using this model - which it does surprisingly well. Thus this paper has a keen aim: it wants to amalgamate religious concepts with concepts of theoretical computer science. It is an attempt to think out-of-the-box, having fun looking at dogmatic things from a total different perspective, breaking down conventional view on old religious things and does not take itself too serious - after all the least thing we need is a new dogma or idealism, the world is full of it.

NOTE: not falsifyable, thus no scientific theory but its a framework within theories/statements which are falsifiable can be formulated

[ ] lucifer was against this simulation as he thought rebellion is the way to go and to change existence
[ ] but love is much more effective in the longrun and even more rebellious than rebellion itself
[ ] also there were always two fundamentally opposing ways in this existence: overthrow the gods and work against them (left handed path) or succumb to it and try to move through it.
[ ] what is open to question, is what darkness is

\bigskip

"In the beginning there was nothing, which exploded" - Terry Pratchett
\end{abstract}

\input{./tex/introduction.tex}

\input{./tex/background.tex}

\input{./tex/realityABS.tex}

\input{./tex/metaphysics.tex}

\input{./tex/discussion.tex}

\input{./tex/conclusion.tex}

\bibliographystyle{acm}
\bibliography{../../../references/phdReferences.bib}

\end{document}