\section{Chapter III: Sex, Culture, And Conflict: The Emergence Of History}

\subsection{Secual Reproduction}
I was unable to reproduce the Figure III-1, it seems completely unrealistic. Also it contradicts Animation III-2, which looks pretty much like my run: we know that the 2d space is 50x50 = 2500 cells. Now we can roughly estimate that in Animation III-2 Picture 6 the 2d space is covered for more than 50\% which would result in at least 1250 Agents alive. This clearly contradicts the Figure III-1. Also when thinking about it Figure III-1 seems not realistic as there will be about as many agents to cover both sugar-hills except the lowest region of level 1. This is due to the fact that this seems to be enough to sustain them and also they wont move much and just harvest whats there thus accumulating ever increasing wealth and thus being able to mate and give birth to new children. Thus my conclusion: a more realistic value is about 1500 Agents which stays pretty constant.

TODO: export Mean vision and Metabolism as Figure III-2

\subsubsection{Regimes of Population Dynamics}
TODO: experiment with the given parameter to see if it replicates

\subsubsection{Gini Effect of Inheritance}
TODO: implement

\subsubsection{Genealogical Networks}
TODO: implement

\subsection{Cultural Processes}
\subsubsection{Cultural Dynamics}
Couldn't replicate the dynamics of the cultural process as in Figure III-8. both tribes seem to deviate around the same fraction. NetLogoCW also doesn't replicate these dynamics.

Sex is explicitly OFF

BirthOnDeath is not specified but it is probably off, because at this point they haven't explained the inheritance of cultural tags yet
Age-Range is not specified but it is probably infinite. Either the Age-Range is infinite or BirthOnDeath is turned on otherwise the agents would simply die out and in the dynamics Figure III-8 they reported to arrive at "all blue" after around 2500 steps

\subsubsection{Vertical Transmission of Culture}
Implemented
No dynamics reported

\subsubsection{Network of Friends}
TODO: implement

\subsection{Combat}