\subsection{1st Year: Groundwork}
In this year I will learn basics and develop and research the methodology I will use for the main work in the 2nd year. Also I want to write a paper about this work and try to publish it on a Conference or Journal. The time-frame will be set by the date of the 1st year annual oral report which will happen beginning of July thus there are about 6 Months time (counting including January '17 and leaving out July '17). These are the things I want / need to achieve this year:

[ ] apply to bilateral trading
[ ] apply to artificial economies (probably too complicated?)
[ ] todo: implement functional reactive abs (FrABS)
[ ] todo: define an EDSL for FrABS which is at the same time model specification- and implementation-language
[ ] todo: define semantics of FrABS based on semantics of FRP and Actors
[ ] what is emergence and can we reason about it? 
[ ] identify emergent properties: equilibrium, behaviour on macroscale not defined on micro, chaos,...
[ ] if yes can we reason about it in advance or only after having run the simulation and identified emergent properties?
[ ] in other words: can emergence be formalized? if yes how far?
[ ] what is the relation between emergence and computation? we need change over time (=computation) for emergence
[ ] baas 1994: emergence, hierarchies and hyperstructures
[ ] introduce MetaABS (also implement it in the EDSL), try to reason about its application to SugarScape and compare results
[ ] until 1st year report (july): basic implementation of FrABS, literature research about emergence, basic study of MetaABS, reasoning about H&C cross patterns, writing 1st year report
[ ] until end of 1st year (sept): fully implement FrABS with EDSL, completely implement sugarscape

\begin{itemize}
\item Prototyping in Haskell, Scala and Java
\item Study Actor-Model theory: Hewitt, Greif, Clinger, Agha
\item Get into reasoning about programs
\item Basics of Economics \cite{bowles_understanding_2005}, \cite{kirman_complex_2010}
\item Basics of ACE: Tesfatsion
\item Implement PureAgents Library
\item Write \& publish papers
\item Write 1st year report
\item Prototype ACE using Akka and Haskell
\end{itemize}

Important and mile-stones:

\begin{center}
\begin{tabular}{ c|c } 
	End of March & Finished and submit Paper \\ 
	\hline
	June (Mid) & Finished writing 1st year report  \\ 
	\hline
	July & Oral annual report \\
\end{tabular}
\end{center}

% JANUARY --------------
\subsubsection{January}
\paragraph{Research} Finish implementation work in Java and Haskell

\paragraph{Paper} First draft
	
\paragraph{Reading}
	\begin{itemize}
		\item Book "Gleichzeitige Ungleichzeitigkeiten"
		\item Complex Systems Stuff
		\item Find semantics of simulation papers
	\end{itemize}
	
% February --------------	
\subsubsection{February}
\paragraph{Research}
	\begin{itemize}
		\item Theoretical: Specification language for ABS in Haskell.
		\item Programming: ABS and Yampa.
	\end{itemize}
		
\paragraph{Paper} Refine paper: read by supervisors and colleagues
	
\paragraph{Reading}
	\begin{itemize}
		\item Re-Read existing papers on functional programming and simulations 
		\item Read Semantics of Actors (Greif \& Clinger)
		\item ACE papers
	\end{itemize}
	
% March --------------	
\subsubsection{March}
\paragraph{Research}
	\begin{itemize}
		\item Theoretical: Specification language for ABS in Haskell.
		\item Programming: ABS and Yampa.
	\end{itemize}
	
\paragraph{Reading}
	\begin{itemize}
		\item Read Semantics of Actors (Greif \& Clinger)
		\item ACE papers
	\end{itemize}
	
\paragraph{Paper} Finalize and submit

% April --------------	
\subsubsection{April}
\paragraph{Research} ABS and Yampa
	
\paragraph{Paper} Write potential conference papers on "specification language" and "reactive agents"
	
\paragraph{Reading} ACE papers

% May --------------	
\subsubsection{May}
\paragraph{Paper} Continue writing potential conference papers on "specification language" and "reactive agents"

\paragraph{Reading} Read papers not read so far but include in 1st year report
	
\paragraph{Writing} Start writing 1st year report

% June --------------	
\subsubsection{June}
\paragraph{Research} Do 1st Year oral exam

\paragraph{Reading} Read papers  not read so far but include in 1st year report

\paragraph{Writing} Finalize 1st year report

% July - September --------------	
\subsubsection{July to September}
\paragraph{Research} Prototyping ACE using my Haskell-Library and Akka: bilateral trading, EDSL for ACE and qualitative modelling, reasoning in ACE
	
\paragraph{Reading}
	\begin{itemize}	
		\item ACE basics papers (Tesfatsion)
		\item ACE models for trading/bartering (e.g. Gintis)
	\end{itemize}
	 
% October --------------	
\subsubsection{October}
2nd year starts