\section{Introduction}
The traditional approach to Agent-Based Simulation (ABS) has so far always been object-oriented techniques, due to the influence of the seminal work of Epstein et al \cite{epstein_growing_1996} in which the authors claim "[..] object-oriented programming to be a particularly natural development environment for Sugarscape specifically and artificial societies generally [..]" (p. 179). This work established the metaphor in the ABS community, that \textit{agents map naturally to objects} \cite{north_managing_2007} which still holds up today.

In this paper we challenge this metaphor and explore ways of approaching ABS using the functional programming paradigm with the language Haskell. We present fundamental concepts and advanced features of functional programming and we show how to leverage the benefits of it \cite{hudak_haskell_1994, hudak_history_2007} to become available when implementing ABS functionally. 

We claim that the community needs functional programming in ABS because of its \textit{scientific computing} nature where results need to be reproducible and correct while simulations should be able to massively scale-up as well. The established object-oriented approaches need considerably high effort and might even fail to deliver these objectives due to its conceptually different approach to computing. In contrast, we claim that by using functional programming for implementing ABS it is easy to add parallelism and concurrency, the resulting simulations are easy to test and verify, guaranteed to be reproducible already at compile-time, have few potential sources of bugs and are ultimately very likely to be correct.

The aim of this paper is to conceptually show \textit{how} to implement ABS in functional programming using Haskell and \textit{why} it is of benefit of doing so. Further, we give the reader a good understanding of what functional programming is, what the challenges are in applying it to ABS and how we solve these in our approach. 

Although functional programming is a highly technical subject, we avoid technical discussions and follow a very high-level approach, focusing on the concepts instead of their implementation. Still, it is necessary to present a bit of code to make things more clear. We always explain the code on a conceptual level, not necessarily being very precise and technical correct in terms of functional programming but here we aim for readability and clarity instead of technical perfection and detail. For readers which are interested in learning functional programming and go into technical details we refer to relevant literature in the respective parts of the paper.

The papers contribution is the introduction of the functional programming paradigm using Haskell to ABS on a \textit{conceptual} level, identifying benefits, difficulties and drawbacks. It shows on a conceptual level that functional programming can be of benefit in ABS to write simulation software which is easier to parallelise and add concurrency, has less sources of bugs, is more likely to be correct and guaranteed to be reproducible already at compile-time. Also this work can be seen as a mini review paper on the topic of using functional programming in ABs, citing the relevant literature in the respective sections. To the best of our knowledge, it is the first one to do so.