\documentclass[format=acmsmall, review=false, screen=true]{acmart}		% ICFP
%\documentclass[format=sigplan, review=true]{acmart}		% HASKELL SYMPOSIUM 
%\documentclass[format=sigconf, review=true]{acmart}		% IFL

%\usepackage{float}
%\usepackage{graphicx}
%\usepackage{subcaption}
%\usepackage{ifthen}
\usepackage{minted}
%\usepackage{verbatim}

% Metadata Information
%% use defaults for review submission.
%\acmConference[HS18]{Haskell Symposium}{2018}{09}
%\acmYear{2018}
%\copyrightyear{2018}
\acmConference[IFL'18]{International Symposium on Implementation and Application of Functional Languages}{August 2019}{Lowell, MA, USA}
\acmYear{2019}
\copyrightyear{2019}
%\acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}

% Copyright
%% use 'none' for review submission.
\setcopyright{none}
%\setcopyright{acmcopyright}	% = copyright transfer to ACM
%\setcopyright{acmlicensed} 		% = retaining copyright but granting ACM exclusive publication rights
%\setcopyright{rightsretained}  % = open access on payment of a fee
%\setcopyright{usgov}
%\setcopyright{usgovmixed}
%\setcopyright{cagov}
%\setcopyright{cagovmixed}

% TODO : get the data
% DOI
% \acmDOI{0000001.0000001}

% TODO: fill in
% Paper history
\received{May 2018}
%\received[revised]{March 2018}
%\received[accepted]{March 2018}

% Document starts
\begin{document}

\newminted[HaskellCode]{haskell}{fontsize=\small}

% Title portion. Note the short title for running heads
\title[The Agent's new Cloths]{The Agent's new Cloths}
\subtitle{Towards functional programming in Agent-Based Simulation}

\author{Jonathan Thaler}
%\orcid{TODO}
\email{jonathan.thaler@nottingham.ac.uk}
\author{Thorsten Altenkirch}
\email{thorsten.altenkirch@nottingham.ac.uk}
\affiliation{%
  \institution{University of Nottingham}
  \streetaddress{7301 Wollaton Rd}
  \city{Nottingham}
  \postcode{NG8 1BB}
  \country{United Kingdom}}

\begin{abstract}
TODO: select journals
- ACM Transactions on Modeling and Computer Simulation (TOMACS): https://tomacs.acm.org/
- Journal of Simulation: https://www.tandfonline.com/toc/tjsm20/current (https://www.tandfonline.com/doi/10.1057/jos.2010.3)
- JASSS: http://jasss.soc.surrey.ac.uk/JASSS.html

TODO: although we approach it from a high level, we have to present a bit of code at various points otherwise its too vague. 

IMPORTANT: don't over-exaggerate: instead of harder, say hard. instead of easier, say easy

TODO: it is paramount not to write against the established approach but for the functional approach. not to try to come up with arguments AGAINST the object-oriented approach but IN FAVOUR for the functional approach. In the end: don't tell the people that what they do sucks and that i am the saviour with my new method but: that i have a new method which might be of interest as it has a few nice advantages.

So far, the pure functional paradigm hasn't got much attention in Agent-Based Simulation (ABS) where the dominant programming paradigm is object-orientation, with Java, Python and C++ being its most prominent representatives. We claim that functional programming using Haskell is very well suited to implement complex, real-world agent-based models and brings with it a number of benefits. In this paper we will introduce the reader to the functional programming paradigm and explain how it can be applied to implementing ABS. Further we discuss benefits and advantages. As use-case we implemented the seminal Sugarscape model in Haskell.
\end{abstract}

%
% The code below should be generated by the tool at
% http://dl.acm.org/ccs.cfm
% Please copy and paste the code instead of the example below.
%
% TODO needs to be generated
%\begin{CCSXML}
%<ccs2012>
% <concept>
%  <concept_id>10010520.10010553.10010562</concept_id>
%  <concept_desc>Computer systems organization~Embedded systems</concept_desc>
%  <concept_significance>500</concept_significance>
% </concept>
% <concept>
%  <concept_id>10010520.10010575.10010755</concept_id>
%  <concept_desc>Computer systems organization~Redundancy</concept_desc>
%  <concept_significance>300</concept_significance>
% </concept>
% <concept>
%  <concept_id>10010520.10010553.10010554</concept_id>
%  <concept_desc>Computer systems organization~Robotics</concept_desc>
%  <concept_significance>100</concept_significance>
% </concept>
% <concept>
%  <concept_id>10003033.10003083.10003095</concept_id>
%  <concept_desc>Networks~Network reliability</concept_desc>
%  <concept_significance>100</concept_significance>
% </concept>
%</ccs2012>
%\end{CCSXML}
%
%\ccsdesc[500]{Computer systems organization~Embedded systems}
%\ccsdesc[300]{Computer systems organization~Redundancy}
%\ccsdesc{Computer systems organization~Robotics}
%\ccsdesc[100]{Networks~Network reliability}

%
% End generated code
%

\keywords{Agent-Based Simulation, Functional Programming, Haskell}

\maketitle

\input{./tex/introduction.tex}

\input{./tex/bugsAbs.tex}

\input{./tex/fpBackground.tex}

\input{./tex/fpAdvanced.tex}

\input{./tex/relatedResearch.tex}

\input{./tex/functionalView.tex}

\input{./tex/multiMethod.tex}

\input{./tex/discussion.tex}

\input{./tex/conclusions.tex}

\input{./tex/furtherResearch.tex}

\begin{acks}
The authors would like to thank
\end{acks}

% Bibliography
\bibliographystyle{ACM-Reference-Format}
%% Citation style
%% Note: author/year citations are required for papers published as an
%% issue of PACMPL.
%%\citestyle{acmauthoryear}   %% For author/year citations
\bibliography{../../../references/phdReferences.bib}

\end{document}
