\documentclass[format=acmsmall, review=true, screen=true]{acmart}		% ICFP
%\documentclass[format=sigplan, review=true]{acmart}		% HASKELL SYMPOSIUM 
%\documentclass[format=sigconf, review=true]{acmart}		% IFL

\usepackage{float}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{ifthen}
\usepackage{minted}
\usepackage{verbatim}

% Metadata Information
%% use defaults for review submission.
%\acmConference[HS18]{Haskell Symposium}{2018}{09}
%\acmYear{2018}
%\copyrightyear{2018}
\acmConference[IFL'18]{International Symposium on Implementation and Application of Functional Languages}{August 2019}{Lowell, MA, USA}
\acmYear{2019}
\copyrightyear{2019}
%\acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}

% Copyright
%% use 'none' for review submission.
\setcopyright{none}
%\setcopyright{acmcopyright}	% = copyright transfer to ACM
%\setcopyright{acmlicensed} 		% = retaining copyright but granting ACM exclusive publication rights
%\setcopyright{rightsretained}  % = open access on payment of a fee
%\setcopyright{usgov}
%\setcopyright{usgovmixed}
%\setcopyright{cagov}
%\setcopyright{cagovmixed}

% TODO : get the data
% DOI
% \acmDOI{0000001.0000001}

% TODO: fill in
% Paper history
\received{May 2018}
%\received[revised]{March 2018}
%\received[accepted]{March 2018}

% Document starts
\begin{document}

\newminted[HaskellCode]{haskell}{fontsize=\footnotesize}

% Title portion. Note the short title for running heads
\title[Implementing Correct-By-Construction System Dynamics]{Implementing Correct-By-Construction System Dynamics}
\subtitle{A pure functional approach}

\author{Jonathan Thaler}
%\orcid{TODO}
\email{jonathan.thaler@nottingham.ac.uk}
%\author{Peer-Olaf Siebers}
%\email{peer-olaf.siebers@nottingham.ac.uk}
%\author{Thorsten Altenkirch}
%\email{thorsten.altenkirch@nottingham.ac.uk}
\affiliation{%
  \institution{University of Nottingham}
  \streetaddress{7301 Wollaton Rd}
  \city{Nottingham}
  \postcode{NG8 1BB}
  \country{United Kingdom}}

\begin{abstract}
%TODO: select journals
%- ACM Transactions on Modeling and Computer Simulation (TOMACS): https://tomacs.acm.org/
%- ICFP 2019 functional pearl?

TODO: it is yet unclear where i will submit this paper thus the audience is not clear. depending on which audience (simulation or functional programming) i have to rephrase the 'problem' and the 'claims' and 'contributions':
	- for the functional programming people can stay as it is but i might need to clarify on System Dynamics
	- for the simulation people i might need more clarification on Haskell and functional programming and rephrase the 'problem' and the 'claims' and 'contributions' to something they can understand easier. Problem is that i must not fall into the trap of selling it as FP vs OOP because there is FRP also in e.g. Java. 
TODO: write related work

In this short paper we investigate how to implement System Dynamics in the functional programming language Haskell. We use the concept of Functional Reactive Programming which allows to express continuous-time systems in a functional way. We show that System Dynamics map naturally to the abstractions provided by Functional Reactive Programming. Together with Haskell's strong static type system, we arrive at a correct-by-construction implementation which deterministic reproducibility we can guarantee at compile-time.
\end{abstract}

%
% The code below should be generated by the tool at
% http://dl.acm.org/ccs.cfm
% Please copy and paste the code instead of the example below.
%
% TODO needs to be generated
%\begin{CCSXML}
%<ccs2012>
% <concept>
%  <concept_id>10010520.10010553.10010562</concept_id>
%  <concept_desc>Computer systems organization~Embedded systems</concept_desc>
%  <concept_significance>500</concept_significance>
% </concept>
% <concept>
%  <concept_id>10010520.10010575.10010755</concept_id>
%  <concept_desc>Computer systems organization~Redundancy</concept_desc>
%  <concept_significance>300</concept_significance>
% </concept>
% <concept>
%  <concept_id>10010520.10010553.10010554</concept_id>
%  <concept_desc>Computer systems organization~Robotics</concept_desc>
%  <concept_significance>100</concept_significance>
% </concept>
% <concept>
%  <concept_id>10003033.10003083.10003095</concept_id>
%  <concept_desc>Networks~Network reliability</concept_desc>
%  <concept_significance>100</concept_significance>
% </concept>
%</ccs2012>
%\end{CCSXML}
%
%\ccsdesc[500]{Computer systems organization~Embedded systems}
%\ccsdesc[300]{Computer systems organization~Redundancy}
%\ccsdesc{Computer systems organization~Robotics}
%\ccsdesc[100]{Networks~Network reliability}

%
% End generated code
%

\keywords{System Dynamics, Functional Reactive Programming, Haskell}

\maketitle

\input{./tex/introduction.tex}

\input{./tex/related.tex}

\input{./tex/background.tex}

\input{./tex/sir.tex}

\input{./tex/implementation.tex}

\input{./tex/discussion.tex}

\input{./tex/conclusions.tex}

\begin{acks}
The authors would like to thank J. Hey for constructive feedback, comments and valuable discussions.
\end{acks}

% Bibliography
\bibliographystyle{ACM-Reference-Format}
%% Citation style
%% Note: author/year citations are required for papers published as an
%% issue of PACMPL.
%%\citestyle{acmauthoryear}   %% For author/year citations
\bibliography{../../../references/phdReferences.bib}

\clearpage

\appendix

\input{./tex/app_code.tex}

\end{document}
