\chapter{Publication Plan and Thesis Structure}

\section{Publication Plan}
\begin{enumerate}
	\item pure functional paper: selling ABS in a functional way to the functional community. conference. proves my methodology: how can we do it, what are the benefits. Also presents very foundations and fundamental concepts of ABS.
	
	\item conceptual paper about functional programming: why do we need functional programming in agent-based simulation? will be the introduction which builds up the argument and story. journal: https://tomacs.acm.org/
	
	\item stm paper: we can do parallel and concurrent ABS in haskell without the low level pain. journal: https://tomacs.acm.org/. showing that functionals strengths in parallelisation and concurrency can be applied to ABS as well.
	
	\item sd in haskell paper: we can do system dynamics in functional programming as well arriving at a correct-by-construction approach. conference e.g. experience paper to ICFP? or tomacs as an introductory paper?
	
	\item testing of agent-based simulations: there doesn't exist much work on testing and verification of agent-based simulation. this paper investigates this on a more technical level using unit- and property-tests in the context of functional agent-based simulation.

	\item sync communication in FRP: how can signal functions communicate with each other synchronously? this means basically that we change the network structure dynamically all the time. in the end its a pure functional way of method calls. Is necessary for an important mechanism required in ABS which needs to be approached different in FP. journal
	
	\item dependent types in ABS: present an indexed monad for ABS which allows correct by construction ABS following a shallow embedded EDSL. conference. showing how we can arrive at software which is correct by construction.
	
	\item totality - equilibrium correspondence: dependent types to bring ABS and SD together and proof that a simulation with an equilibrium is total: journal. 
\end{enumerate}