%\documentclass[a4paper, 10pt, conference]{../../templates/IEEEconf/IEEEconf}
\documentclass[10pt, conference, onecolumn]{../../../templates/IEEEtran/IEEEtran}
%\documentclass[10pt, journal]{../../templates/IEEEtran/IEEEtran}

\usepackage{graphicx}
\usepackage{caption} 
\usepackage{subcaption}
\usepackage{hyperref}
\usepackage{listings}
\usepackage{hhline}
\usepackage{float}
\usepackage{amssymb}
\usepackage[autostyle=true]{csquotes}
\usepackage{amsmath}
\usepackage{marvosym}
\usepackage{minted}

\font\subtitlefont=cmr12 at 18pt

% First Contact
\title{Spreading the disease \\ {\huge Simulating epidemics using Functional Reactive Programming} \\ {\large An Agent-Based Approach}}

% IEEEtran journal authors
%\author{Jonathan Thaler, ̃Peer-Olaf Siebers \\ School of Computer Science \\ University of Nottingham%
%\thanks{jonathan.thaler@nottingham.ac.uk}%
%\thanks{peer-olaf.siebers@nottingham.ac.uk}
%}

% IEEEtran conference authors
\author{
	\IEEEauthorblockN{Jonathan Thaler}
	\IEEEauthorblockA{School of Computer Science\\
		University of Nottingham\\
		jonathan.thaler@nottingham.ac.uk}
		
	\and
		
	\IEEEauthorblockN{Thorsten Altenkirch}
	\IEEEauthorblockA{School of Computer Science\\
		University of Nottingham\\
		thorsten.altenkirch@nottingham.ac.uk}
}

%\IEEEpubid{0000--0000/00\$00.00 ̃\copyright ̃2015 IEEE}

% IEEEconf authors
%\author{
%	Jonathan Thaler \\
%	\email{jonathan.thaler@nottingham.ac.uk} \\
%	\begin{affiliation}
%		School of Computer Science, University of Nottingham
%	\end{affiliation} \\
%	\and 
%	Peer-Olaf Siebers \\
%	\email{peer-olaf.siebers@nottingham.ac.uk} \\
%	\begin{affiliation}
%		School of Computer Science, University of Nottingham
%	\end{affiliation} 
%	\and 
%	Thorsten Altenkirch \\
%	\email{thorsten.altenkirch@nottingham.ac.uk} \\
%	\begin{affiliation}
%		School of Computer Science, University of Nottingham
%	\end{affiliation} 
%}

\begin{document}
\maketitle 

\begin{abstract}
TODO: feedback by thorsten: its too hardcore! instead of presenting my library we need to do "how to arrive at FrABS from Haskell to Yampa to FrABS" in which we (thorsten and me) derive step-by-step an implementation of a very simple agent-based model using Haskell and Yampa. The conclusion will then be that we provide all stuff already in FrABS. 
TODO: split the approach to FrABS into the following steps
	1.: run in random monad without any reactive features. problem: need to drag time & state around
	2.: run in Yampa with reactive features. problem: randomness is tedious
	3.: run in Dunai.
	
TODO: need to select the right journal for publication, something more practical functional programming

TODO: pictures should show up where they are mentioned in the text

TODO: cite my own 1st paper from SSC2017: add it to citations

TODO: refine it: start with simulating epidemics and then go into ABS

Agent-Based Simulation (ABS) is a methodology in which a system is simulated in a bottom-up approach by modelling the micro interactions of its constituting parts, called agents, out of which the global macro system behaviour emerges. So far, the Haskell community hasn't been much in contact with the community of ABS due to the latter's primary focus on the object-oriented programming paradigm. This paper tries to bridge the gap between those two communities by introducing the Haskell community to the concepts of ABS. We do this by deriving an agent-based implementation for the simple SIR model from epidemiology. In our approach we leverage the basic concepts of ABS with functional reactive programming from Yampa which results in a surprisingly fresh, powerful and convenient EDSL for formulating ABS in Haskell.
\end{abstract}

\begin{IEEEkeywords}
Functional Reactive Programming, Agent-Based Simulation
\end{IEEEkeywords}

\input{./tex/introduction.tex}

\input{./tex/definingABS.tex}

\input{./tex/sirModel.tex}

\input{./tex/functionalReactiveSIR.tex}

\input{./tex/nonReactive.tex}

\input{./tex/execRepls.tex}

\input{./../shared/tex/results.tex}

\input{./tex/spatiality.tex}

\input{./tex/superSamplingRandomness.tex}

\input{./tex/agentSignals.tex}

\input{./tex/discussion.tex}

\input{./tex/relatedResearch.tex}

\input{./tex/conclusion.tex}

\input{./../shared/tex/acknowledgments.tex}

\bibliographystyle{../../../templates/IEEEtran/bibtex/IEEEtran}
\bibliography{../../../../references/phdReferences.bib}

\appendices

\newpage
\input{./../shared/tex/absCode.tex}

\end{document}