\documentclass{article}

\usepackage{hyperref}
\usepackage[utf8]{inputenc}
\usepackage{graphicx} % Required for the inclusion of images
\usepackage{amsmath} % Required for some math elements 
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\newtheorem{theorem}{Theorem}
\newtheorem{corollary}{Corollary}[theorem]
\newtheorem{lemma}[theorem]{Lemma}

%TODO follow the instructions on http://workspace.nottingham.ac.uk/display/CompSci/Procedure+for+PhD+Annual+Reviews

% A written report that contains the following sections. There is no specific requirement on formatting but it should be well-presented and proof-read to ensure it is a piece of high-quality academic writing. As guidance, the whole report including the appendix should normally be around 8000-10000 words in length.

%TODO Current title of PhD project, Whether it is a first, second or third annual review report, word count, student name, student ID, names of supervisors, initial registration date, expected date for entering thesis pending stage, expected date for thesis submission.
\title{1st Year Report \\ Pure Functional Methods in \\ Agent-Based Modelling \& Simulation} % Title

\author{Jonathan \textsc{Thaler} \\ jonathan.thaler@nottingham.ac.uk} % Author name

\date{\today} % Date for the report

\begin{document}
\maketitle % Insert the title, author and date

% If you wish to include an abstract, uncomment the lines below
\begin{abstract}
A succinct and concise summary (250 words maximum) of the report contents and presented on a single page.

So far specifying Agent-Based Models and implementing them as an Agent-Based Simulation (ABS) is done using object-oriented methods like UML and object-oriented programming languages like Java. The reason for this is that until now the concept of an agent was always understood to be very close to, if not equals to - which it is not - the concept of an object. Therefore, the reasoning goes, object-oriented methods and languages should fit themselves naturally to specify and implement agent-based simulations. In this PhD Thesis we fundamentally challenge this assumption by investigating how Agent-Based Models and Simulations can be specified and implemented using pure functional methods and programming in Haskell and what the benefits are. We will show that the implicit assumption that an Agent is \textit{about equal to} an Object is not correct and leads to many implicit assumptions in object-oriented implementations of Agent-Based Simulation (ABS). When implementing ABS in Haskell these implicit assumptions become explicit and challenge the fundamental assumptions about ABS and Agents. We present these implicit assumption in an explicit way by approaching it through programming, type-theory and category-theory to further deepen the concepts and methods in the field of Agent-Based Modelling \& Simulation. We also think that the major benefit of implementing ABS in Haskell is the potential for an unprecedented approach of formal validation \& verification of an Agent-Based Model and its implementation. Due to the declarative nature of pure functional programming in Haskell it is possible to implement an EDSL for ABS which ideally results in code which looks like specification thus closing the gap between specification and implementation because the specification is already the code. For validation we want to pursue testing through QuickCheck.

In this report I discuss the research conducted to far, present the open problems together with an in-depth literature review. An outline for the research of the following 2 years is given and the aims.
\end{abstract}

%TODO Contents Page Main headings with page numbers.

\input{./tex/introduction.tex}

\input{./tex/literatureReview.tex}

\input{./tex/aimsObjectives.tex}

\input{./tex/workToDate.tex}

\input{./tex/conclusions.tex}

\input{./tex/futureWork.tex}

%\bibliographystyle{acm}
\bibliography{../../references/phdReferences}

\input{./tex/appendix.tex}

\end{document}
