\section{Deriving a pure functional approach}
\label{sec:functional_approach}

%2 main points from Henrik: 1.: try to stick to synchronous updates because this is how the real world works. 2.: get rid of globally shared mutable state as it complicates things extremely with reasoning
%clear conceptual formal model of what agents are and then structure my implementation around it

% TODO: put explanations before the code, otherwise readers get confused e.g. stepSimulation function

We presented a high-level agent-based approach to the SIR model in the previous section, which focused only on the states and the transitions, but we haven't talked about technical implementation details on how to actually implement such a state-machine. The authors of \cite{thaler_art_2017} discuss two fundamental problems of implementing an agent-based simulation from a programming language agnostic point of view. The first problem is how agents can be pro-active and the second how interactions and communication between agents can happen. For agents to be pro-active they must be able to perceive the passing of time, which means there must be a concept of an agent-process which executes over time. Interactions between agents can be reduced to the problem of how an agent can expose information about its internal state which can be perceived by other agents. \\
In this section we will derive a pure functional approach for an agent-based simulation of the SIR model in which we will pose solutions to the previously mentiond problems. We will start out with a very naive approach and show its limitations which we overcome by adding FRP. Then in further steps we will add more concepts and generalisations, ending up at the final approach which utilises monadic stream functions (MSF), a generalisation of FRP 
\footnote{The code of all steps can be accessed freely through the following URL: \url{https://github.com/thalerjonathan/phd/tree/master/public/purefunctionalepidemics/code}}.

% TODO: put all functions in italics

\input{./tex/step1.tex}

\input{./tex/step2.tex}

\input{./tex/step3.tex}

\input{./tex/step4.tex}

\input{./tex/step5.tex}

% this is omited for now as it takes too much time and involves too many details but would need much more explanation still
%\input{./tex/step6.tex}\\

% we have also a step7: event-driven implemented but omit it due to lack of space

\subsection{Further Steps}
\subsubsection{Agent-Transactions}
Agent-transactions are necessary when an arbitrary number of interactions between two agents need to happen instantaneously without time-lag. The use-case for this are price negotiations between multiple agents where each pair of agents needs to come to an agreement in the same time-step \cite{epstein_growing_1996}. In object-oriented programming, the concept of synchronous communication between agents is implemented directly with method calls. We have implemented synchronous interactions, which we termed agent-transactions in an additional step which we had to omit due to lack of space. We solved it pure functionally by running the signal functions of the transacting agent pair as often as their protocol requires but with $\Delta t=0$, which indicates the instantaneous character of agent-transactions.

\subsubsection{Event Scheduling}
Our approach is inherently time-driven where the system is sampled with fixed $\Delta t$. The other fundamental way to implement an ABS in general, is to follow an event-driven approach \cite{meyer_event-driven_2014} which is based on the theory of discrete-event simulation \cite{zeigler_theory_2000}. In such an approach the system is not sampled in fixed $\Delta t$ but advanced as events occur where the system stays constant in between. Depending on the model, in an event-driven approach it may be more natural to express the requirements of the model. In an additional step we have implemented a rudimentary event-driven approach which allows the scheduling of events but had to omit it due to lack of space. Using the flexibility of MSFs we added a State transformer to the monad stack which allows enqueuing of events into a priority queue. The simulation is advanced by processing the next event at the top of the queue which means running the MSF of the agent which receives the event. The simulation terminates if there are either no more events in the queue or after a given number of events or if the simulation time has advanced to some limit. Having made the transition to MSFs, implementing this feature was quite easy which shows the power and strength of the generalised approach to FRP using MSFs.

\subsubsection{Dynamic Agent creation}
In the SIR model, the agent population stays constant - agents don't die and no agents are created during simulation - but some simulations \cite{epstein_growing_1996} require dynamic agent creation and destruction. We can easily add and remove agents signal functions in the recursive switch after each time-step. The only problem is that creating new agents requires unique agent ids but with the transition to MSFs we can add a monadic context which allows agents to draw the next unique agent id when they create a new agent. %The id generation process should happen in the agent as the creating agent almost always needs then to communicate with this new agent. If we defer the id generation to the simulation system itself then we need a mechanism to feed it back to the creating agent, which could become quite cumbersome.
