\section{Conclusions}
\label{sec:conclusions}
The case-study strongly hints that our claim that pure FP has indeed its place in ABS is valid but we conclude that it is yet too early to pick up this paradigm for ABS. We think that engineering a proper implementation of a complex ABS model takes substantial effort in pure FP due to different techniques required. We believe that at the moment such an effort pays off only in cases of high-impact and large-scale simulations which results might have far-reaching consequences e.g. influence policy decisions. It is only there where the high requirements for reproducibility, robustness and correctness provided by FP are really needed. Still, we plan on distilling the developed techniques of the case-study into a general purpose ABS library. This should allow implementing models much easier and quicker, making the pure FP approach an attractive alternative for prototyping, opening the direction for a broad use of FP in the field of ABS.

\subsection{Further Research}
Often, ABS models build on an underlying DES core, especially when they follow an event-driven approach \cite{meyer_event-driven_2014} where they need to schedule actions into the future. Due to the time-driven approach of the Sugarscape model, we didn't implement this in our case-study but our pure functional simulation core is conceptually already very close to a DES approach. We plan on extending it to allow true event-based ABS, which we want to feed into our general purpose ABS library as well.

% POTENTIAL FOR SHORTENING
Due to the recursive nature of FP we believe that it is also a natural fit to implement recursive simulations as the one discussed in \cite{gilmer_recursive_2000}. Recursive ABS gives each Agent the ability to run the simulation locally from its point of view to project its actions into the future and change them in the present. Due to controlled side-effects and referential transparency, combined with the recursive nature of pure FP, we think that implementing a recursive simulation in such a setting should be straight-forward.

This research is only a first step towards an even more extensive use of pure FP systems in ABS. The next step is to investigate the use of dependent types in ABS, using the pure functional language Idris \cite{brady_idris_2013}. Dependent types allow to express even stronger guarantees at compile time, going as far as programs being valid proofs for the correctness of the implemented feature.