\section{The evolution of my computer science interests}
\label{app:cs_interests}

This is a retrospective of how my interests in computer science evolved into the current state. \bigskip

I was and am no genius. I was also not extremely smart, sometimes I was slow in understanding, I was good enough but what set me apart was: if I understood something, I've understood fundamental and I could transfer it to other areas as well. Also if I really wanted to do something I pursued it so long until I could do it (gameengine). And the most important \\
There was always a bit of playfulness in my approaches: it was the mere curiosity how things would unfold, the fun in exploring unknown territory - when I had the feeling I understood enough I moved on to new territories. Thus the major shift of interests from starting programming with age of 15 until now with age of 33 from basic text games to GUI programming to 3D graphics programming with OpenGL 4.0 to web 2.0 development to enterprise back-end applications to operating systems to  embedded systems to system simulation and now to the very basics and mysteries of computation itself.

\bigskip



\subsection{Ethereum}
Based upon a hint by a good friend yesterday (2016 August 17th) I started exploring the idea behind Ethereum which immediately fascinated me. This is a topic I will definitely pursue as a private interest during my PhD Studies. I hope to be able to participate in a project or have the idea for one myself and maybe it can be combined with my PhD Studies (agents running in Ethereum?).