\chapter{Parallelism in ABS}
\label{ch:parallelism_ABS}
The promise of parallelism in Haskell is compelling: speeding up the execution but still retaining all static compile time guarantees about determinism. In other words, using parallelism could give us substantial performance improvement without sacrificing the static guarantees of reproducible outputs from repeated runs with initial conditions.

Generally, parallelism can be applied whenever the execution of code is order-independent, referential transparent, and has no implicit or explicit side effects. In this section we introduce the two most important parallelism concepts of Haskell, \textit{evaluation} and \textit{data-flow} parallelism. Then we discuss their potential use in pure functional ABS in general. We follow \cite{marlow_parallel_2013} and refer to it for an in-depth discussion. Furthermore, we show how these concepts can be added to our previously discussed use cases in Chapters \ref{sec:timedriven_firststep}, \ref{sec:adding_env} and Sugarscape. Whe then compare their performance over the original sequential approaches.

\input{./tex/research/parallelabs/parallelism/evaluation.tex}

\input{./tex/research/parallelabs/parallelism/dataflow.tex}

% IGNORE FOR NOW
%\subsection{Data-structure parallelism}
%An environment could be organised and accessed through such a data-structure, which could potentially lead to big speed ups. Agents could locally read the data-structure data-parallel and the simulation kernel could feed the output of the agents data-parallel back into this structure.
%
%%https://learning.oreilly.com/library/view/parallel-and-concurrent/9781449335939/ch05.html
%
%general solution we opt for is  to run agents in parallel in our approaches. in other abs models we could apply data-structure parallelism and/or data-flow parallelism with huge Performance potential but thats always highly model dependent thus we dont go in depth here
\input{./tex/research/parallelabs/parallelism/casestudies.tex}

\input{./tex/research/parallelabs/parallelism/parallelruns.tex}

\input{./tex/research/parallelabs/parallelism/discussion.tex}